---
title: Introduction to R and RStudio
subtitle: Instituto Nacional de Medicina Genomica
author: __Daniel Rivas__ & __Juan Manuel Mejia Arangure__
date: August 18, 2023
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# #If you want to knit the whole notebook, please specify below correct path to the Rmd-file
knitr::opts_knit$set("/home/nikolay-oskolkov/TARGETWISE/Teaching/R_course_2026/RcourseUNAM/slides/")
```


## Outline - R Programming Fundamentals

- Data structures
- Common operations
  - Control Flow
  - Loops
  - Functions
- Statistics

# Data structures

## Primitives

**Logical** => TRUE(T) | FALSE(F)

**Numeric** => 357, 3.141596, 0, -59

**Character** => 'char', "char"

**Integer** => 9L

**Complex** => 46+5i

## Primitives

**Vectors** -> one-dimensional collection of elements

**Matrices** -> two-dimensional collection of elements (size must be predefined)

**Arrays** -> _n_-dimensional collection of elements

## Primitives

**Factors** -> one-dimensional collection of elements with predefined possible values

**Lists** -> arbitrary collection of any elements

**Data Frames** -> flexible two-dimensional representation of data

# Workshop III

# Common operations

## Control Flow

Problem

## Control Flow

Query data

Decision making

```{r}
# # if statement
# if (condition) {
#   statement
# } else {
#   statement
# }
```

## Loops

Problem

## Loops

Defined or conditional repetition

Iterative

```{r}
# # for loop
# for (value in vector) {
#    statements
# }
```

## Functions

Problem

## Functions

Customized work flow with variable inputs

```{r}
# # no arguments
# funcx <- function() {
#   print('I am a function')
# }

# # arguments
# funcy <- function(x) {
#   x * 10
# }
```

# Workshop IV

# Statistics

## Examples

Mean
```{r}
# mean(x, trim = 0, na.rm = FALSE, ...)
```

Median
```{r}
# median(x, na.rm = FALSE)
```

Summary
```{r}
# summary(object, ...)
```

## Examples

**linear regression** => y = ax + b
```{r}
# lm(formula, data)
```

**multiple regression** => y = a + x1 + x2 + x3 ...
```{r}
# lm(y ~ x1 + x2 + x3..., data)
```

**logistic regression**
```{r}
# glm(formula, data, family)
```

## Examples

**ancova analysis**
```{r}
# aov(formula, data = NULL, projections = FALSE, qr = TRUE, contrasts = NULL, ...)
```

**time series analysis**
```{r}
# survfit(formula)
```

**chi square**
```{r}
# chisq.test(data)
```

# Workshop V

# Questions & Comments

<div class = 'centered'>

Daniel Rivas

danielrivasmd@gmail.com

</div>
