---
title: Introduction to R and RStudio
subtitle: Instituto Nacional de Medicina Genomica
author: __Daniel Rivas__ & __Juan Manuel Mejia Arangure__
date: August 25, 2023
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# #If you want to knit the whole notebook, please specify below correct path to the Rmd-file
knitr::opts_knit$set("/home/nikolay-oskolkov/TARGETWISE/Teaching/R_course_2026/RcourseUNAM/slides/")
```


```{r, echo = F}
library(ggplot2)
```

## Outline - Working with Data

- Working with Files
- Advanced Graphics

# Working with Files

## Data manipulation

- `tidyr` - https://tidyr.tidyverse.org

- `dplyr` - https://dplyr.tidyverse.org

- `data.table` - https://r-datatable.com

## Data import

- `readr` - https://readr.tidyverse.org

- `readxl` - https://readxl.tidyverse.org

- `haven` - https://haven.tidyverse.org

# Advanced Graphics

## Graphics

```{r, echo = F}
set.seed(1919)
x <- rnorm(1000)
y <- x + rnorm(1000)
plot(x, y)
```

## Graphics

```{r, echo = F}
plot(x, y)
abline(lm(y ~ x), col = 'red')
```

## Graphics

```{r, echo = F}
plot(density(x))
```

## Graphics

```{r, echo = F}
plot(density(x))
lines(density(y), col = 'red')
```

## Graphics

```{r, echo = F}
plot(x, y, main = 'This is my Plot', xlab = 'X-Values', ylab = 'Y-Values')
```

## Graphics

```{r, echo = F}
group <- rbinom(1000, 1, 0.3) + 1
plot(x, y, col = group, pch = group)
```

## Graphics

```{r, echo = F}
z <- seq(- 5, 5, 0.01)
plot(z, cos(z))
```

## Graphics

```{r, echo = F}
set.seed(11)
tmp <- sample(1:10, 5, replace = T)
barplot(tmp)
```

## Graphics

```{r, echo = F}
hist(tmp)
```

## Graphics

```{r, echo = F}
et <- c(H = 600, F = 300, C = 150, E = 100, O = 200)
pie(et)
```

## Graphics

```{r, echo = F}
boxplot(airquality$Ozone)
```

## Graphics

```{r, echo = F}
oz <- airquality$Ozone
t <- airquality$Temp
oznorm <- rnorm(200, mean = mean(oz, na.rm = T), sd = sd(oz, na.rm = T))
tnorm <- rnorm(200, mean = mean(t, na.rm = T), sd = sd(t, na.rm = T))
boxplot(oz, oznorm, t, tnorm, at = c(1,2,4,5), names = c('ozone', 'normal', 'temp', 'normal'), las = 2, col = c('orange','red'), border = 'brown', horizontal = T, notch = T)
```

## Graphics

```{r, echo = F}
boxplot(Temp ~ Month, data = airquality, col = 'orange', border = 'brown')
```

## Graphics

- Pie chart

- Bar chart

- Line chart

- Scatterplot

## Graphics

Decompose graphics into its constituents

- theme

- coordinates

- facets

- geometries

- scales

- statistics

- mapping

- data

## Graphics

- `ggplot2`

```{r, eval = F}
ggplot(
  data = faithful,
  mapping = aes(x = eruptions, y = waiting)
) +
  geom_point()
```

Which columns to use for `x` and `y`

How to draw the plot

`+` is used to combine `ggplot2` elements

## Graphics

```{r, echo = T, eval = F}
ggplot(
  data = faithful,
  mapping = aes(x = eruptions, y = waiting)
) +
  geom_point()

ggplot(data = faithful) +
  geom_point(
    mapping = aes(x = eruptions, y = waiting)
  )

ggplot() +
  geom_point(
    mapping = aes(x = eruptions, y = waiting),
    data = faithful
  )
```

## Graphics

```{r, echo = F}
ggplot(faithful) + geom_point(aes(x = eruptions, y = waiting, colour = eruptions < 3))
```

Mapping color

## Graphics

```{r, echo = F}
ggplot(faithful) + geom_point(aes(x = eruptions, y = waiting), colour = 'steelblue')
```

Setting color

## Graphics

```{r, echo = F, message = F}
ggplot(faithful) + geom_histogram(aes(x = eruptions))
```

There are many types of `geoms`, but their mapping requirements differ

## Graphics

```{r, echo = F}
ggplot(faithful, aes(x = eruptions, y = waiting)) + geom_density_2d() + geom_point()
```

Layers are stacked in the order of code appearance

## Graphics

```{r, echo = F}
ggplot(mpg) + geom_point(aes(x = displ, y = hwy)) + facet_wrap(~ class)
```

Faceting is often the best way to avoid overplotting

## Graphics

```{r, echo = F}
ggplot(mpg) + geom_point(aes(x = displ, y = hwy)) + facet_grid(year ~ drv)
```

`facet_grid()` provides a way of doing graphic pivots

## Beyond _ggplot2_

`ggplot2` is huge!

Maintenance is already a team effort

- 47 `geom`

- 25 stats

- 62 scales

Many extensions niche specific & better developed by filed experts

## R Gallery

https://r-graph-gallery.com/

## _ggstatsplot_

https://github.com/IndrajeetPatil/ggstatsplot

# Questions & Comments

<div class = 'centered'>

Daniel Rivas

danielrivasmd@gmail.com

</div>
